Revealing transient strain in geodetic data with Gaussian process regression

### 2017-11-12 RevisedSubmission/hines_hetland_2017_gji.tex
added more

Transient strain rates derived from GNSS data can be used to detect
and understand geophysical processes such as slow slip events and
postseismic deformation. Here we propose using Gaussian process
regression (GPR) as a tool for estimating transient strain rates from
GNSS data. GPR is a non-parametric, Bayesian method for interpolating
scattered data. In our approach, we assume a stochastic prior model
for transient displacements. The prior describes how much we expect
transient displacements to covary spatially and temporally. A
posterior estimate of transient strain rates is obtained by
differentiating the posterior displacements, which are formed by
conditioning the prior with the GNSS data. As a demonstration, we use
GPR to detect transient strain resulting from slow slip events in the
Pacific Northwest. Maximum likelihood methods are used to constrain a
prior model for transient displacements in this region. The temporal
covariance of our prior model is described by a compact Wendland
covariance function, which significantly reduces the computational
burden that can be associated with GPR. Our results reveal the spatial
and temporal evolution of strain from slow slip events. We verify that
the transient strain estimated with GPR is in fact geophysical signal
by comparing it to the seismic tremor that is associated with Pacific
Northwest slow slip events.

@@

Crustal strain rates are fundamentally important quantities for
assessing seismic hazard. Knowing where and how quickly strain is
accumulating gives insight into where we can expect stored elastic
energy to be released seismically. Estimates of secular, or long-term,
strain rates have been used to constrain seismic hazard models such as
UCERF3 \citep{Field2014}. Transient crustal strain, which is caused by
geophysical processes such as slow slip events (SSEs) or postseismic
deformation, is also relevant for assessing seismic hazard. While
transient strain itself is not damaging, there is a possibility that it can
trigger major earthquakes \citep{Roeloffs2006,Freed2001}.

@@

Displacement data from dense GNSS networks, such as the Plate Boundary
Observatory (PBO), can be used to infer secular and transient strain.
Numerous methods for converting GNSS displacement data to strain have
been proposed, and most involve assuming some parametric form for the
deformation signal. The simplest method for estimating secular strain
rates assumes that GNSS derived velocities can be described with a
first-order polynomial (i.e, the deformation gradients are constant)
over some subnetwork of the GNSS stations \citep[e.g.,][]{Feigl1990,
Murray2000}. The components of the secular strain rate tensor for each
subnetwork are then determined from the least squares fit to the
observations. The assumption that deformation gradients are spatially
uniform is not appropriate when subnetworks span a very large area. To
help overcome this deficiency, \citet{Shen1996,Shen2015} used an
inverse distance weighting scheme, in which the estimated secular
strain rate at a point is primarily controlled by observations at
nearby stations. The method of \citet{Shen1996,Shen2015} can be viewed
as a form of local least squares regression with a first-order
polynomial. Other methods for estimating secular strain rates have
parameterized GNSS derived velocities with bi-cubic splines
\citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and
elastostatic Green's functions \citep{Sandwell2016}. The type of basis
functions and the number of degrees of freedom for a parameterization,
which are often chosen subjectively, have a strong influence on the
solution. If there are too few degrees of freedom in the
parameterization, then estimated secular strain rates will be biased
and the uncertainties will be underestimated. On the other hand, if
there are too many degrees of freedom, then there will not be any
coherent features in the estimated secular strain rates. The methods
described by \citet{Beavan2001} and \citet{Tape2009} also require the
user to specify penalty parameters that control a similar trade-off
between bias and variance in the solution. One could also parameterize
deformation with a physically motivated model of interseismic
deformation \citep[e.g.,][]{Meade2005, McCaffrey2007}. In such models
the lithospheric rheology and fault geometries are assumed to be
known. Any errors in the assumed physical model could result in biased
estimates of the secular strain rates and underestimated
uncertainties.

@@

The aforementioned studies are concerned with estimating secular
strain rates. In recent years the Southern California Earthquake
Center (SCEC) community has shown interest in developing methods for
detecting transient deformation. SCEC supported a transient detection
exercise \citep{Lohman2013}, where several research groups tested
their methods for detecting transient geophysical signal with a
synthetic GNSS dataset. Among the methods tested were the Network
Strain Filter (NSF) \citep{Ohtani2010} and the Network Inversion
Filter (NIF) \citep{Segall1997}. The NSF uses a wavelet
parameterization to describe the spatial component of geophysical
signal. The NIF, which is intended for imaging slow fault slip from
geodetic data, uses the elastic dislocation Green's functions from
\citet{Okada1992}. For the NSF and NIF, the time dependence of the
geophysical signal is modeled as integrated Brownian motion. The
method described in \citet{Holt2013} was also tested in the SCEC
transient detection exercise. \citet{Holt2013} detect transient
signal from an estimate of strain rates, and they estimate strain
rates using a bi-cubic spatial parameterization of displacements
between time epochs. \citet{Holt2013} define a detection threshold
based on the strain rate magnitude, and we demonstrate that this is
indeed an effective criterion for identifying a geophysical signal.
For the same reasons described above, the transient deformation and
corresponding uncertainties estimated by these methods can be biased
by the chosen spatial parameterization. It is then difficult to
distinguish signal from noise with these methods, which limits their
utility for transient detection.

@@

Our outlier detection algorithm is loosely based on the data editing
algorithm from \citet{Acheson1975}. Let $\data^*$ denote all $n$ GNSS
displacement observations for a single directional component, which
have been made at positions and times $\points$.  We describe
$\data^*$ as a realization of the random vector
\begin{equation}\label{eq:OutlierDataVec}
\data = \G\mitbf{m} + v(\points) + \mitbf{w},
\end{equation}
where $\G$ and $\mitbf{m}$ are the same as in eq. (\ref{eq:DataVec}),
$v$ is a Gaussian process distributed as $\mathcal{GP}(0,C_v)$, and
$\mitbf{w}$ is a vector of uncorrelated Gaussian noise with known
standard deviations $\mitbf{\sigma} = [\sigma_1, \sigma_2, ...,
\sigma_n]$. The Gaussian process $v$ is intended to describe transient
features in the data that cannot be explained by the linear trend or
seasonal terms in $\G$. We let the temporal covariance of $v$ be a
squared exponential, and we let $v$ be spatially uncorrelated so that,
\begin{equation}
C_v(p,p') = \phi^2\exp\left(\frac{-|t - t'|^2}{2\tau^2}\right) \delta_{\pos,\pos'},
\end{equation}
where $\delta_{\pos,\pos'} = 1$ if $\pos = \pos'$ and $0$ otherwise.
The spatial covariance of $v$ has little effect on the detected
outliers, and so we have assumed that $v$ is spatially uncorrelated
for simplicity. Based on our experience, $v$ can reasonably describe
most transient features in the data when we set $\phi = 1$ mm and
$\tau = 10$ days.

>>>>

Transient strain derived from GNSS data can be used to detect and
understand geophysical processes such as slow slip events and
postseismic deformation. Here we propose using Gaussian process
regression (GPR) as a tool for estimating transient strain from GNSS
data. GPR is a non-parametric, Bayesian method for interpolating
scattered data. In our approach, we assume a stochastic prior model
for transient displacements. The prior describes how much we expect
transient displacements to covary spatially and temporally. A
posterior estimate of transient strain is obtained by differentiating
the posterior transient displacements, which are formed by
conditioning the prior with the GNSS data. As a demonstration, we use
GPR to detect transient strain resulting from slow slip events in the
Pacific Northwest. Maximum likelihood methods are used to constrain a
prior model for transient displacements in this region. The temporal
covariance of our prior model is described by a compact Wendland
covariance function, which significantly reduces the computational
burden that can be associated with GPR. Our results reveal the spatial
and temporal evolution of strain from slow slip events. We verify that
the transient strain estimated with GPR is in fact geophysical signal
by comparing it to the seismic tremor that is associated with Pacific
Northwest slow slip events.

@@

Crustal strain rates are fundamentally important quantities for
assessing seismic hazard. Knowing where and how quickly strain is
accumulating gives insight into where we can expect stored elastic
energy to be released seismically. Estimates of secular, or long-term,
strain rates have been used to constrain seismic hazard models such as
UCERF3 \citep{Field2014}. Transient crustal strain, which is anomalous
strain caused by geophysical processes such as slow slip events (SSEs)
or postseismic deformation, is also relevant for assessing seismic
hazard. While transient strain itself is not damaging, there is a
possibility that it can trigger major earthquakes
\citep{Roeloffs2006,Freed2001}.

@@

Displacement data from dense GNSS networks, such as the Plate Boundary
Observatory (PBO), can be used to infer secular and transient strain.
Numerous methods for converting GNSS displacement data to strain have
been proposed, and most involve assuming some parametric form for the
deformation signal. The simplest method for estimating secular strain
rates assumes that GNSS derived velocities can be described with a
first-order polynomial (i.e, the deformation gradients are constant)
over some subnetwork of the GNSS stations \citep[e.g.,][]{Feigl1990,
Murray2000}. The components of the secular strain rate tensor for each
subnetwork are then determined from the least squares fit to the
velocities. The assumption that deformation gradients are spatially
uniform is not appropriate when subnetworks span a very large area. To
help overcome this deficiency, \citet{Shen1996,Shen2015} used an
inverse distance weighting scheme, in which the estimated secular
strain rate at any point is primarily controlled by the velocities at
nearby stations. The method of \citet{Shen1996,Shen2015} can be viewed
as a form of local least squares regression with a first-order
polynomial. Other methods for estimating secular strain rates have
parameterized GNSS derived velocities with bi-cubic splines
\citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and
elastostatic Green's functions \citep{Sandwell2016}. The type of basis
functions and the number of degrees of freedom for a parameterization,
which are often chosen subjectively, have a strong influence on the
solution. If there are too few degrees of freedom in the
parameterization, then estimated secular strain rates will be biased
and the uncertainties will be underestimated. On the other hand, if
there are too many degrees of freedom, then there will not be any
coherent features in the estimated secular strain rates. The methods
described by \citet{Beavan2001} and \citet{Tape2009} also require the
user to specify penalty parameters that control a similar trade-off
between bias and variance in the solution. One could also parameterize
deformation with a physically motivated model of interseismic
deformation \citep[e.g.,][]{Meade2005, McCaffrey2007}. In such models,
the lithospheric rheology and fault geometries are assumed to be
known. Any errors in the assumed physical model could result in biased
estimates of the secular strain rates and underestimated
uncertainties.

@@

The aforementioned studies are concerned with estimating secular
strain rates. In recent years the Southern California Earthquake
Center (SCEC) community has shown interest in developing methods for
detecting transient deformation. SCEC supported a transient detection
exercise \citep{Lohman2013}, where several research groups tested
their methods for detecting transient geophysical signal with a
synthetic GNSS dataset. Among the methods tested were the Network
Strain Filter (NSF) \citep{Ohtani2010} and the Network Inversion
Filter (NIF) \citep{Segall1997}. The NSF uses a wavelet
parameterization to describe the spatial component of geophysical
signal. The NIF, which is intended for imaging slow fault slip from
geodetic data, uses the elastic dislocation Green's functions from
\citet{Okada1992}. For the NSF and NIF, the time dependence of the
geophysical signal is modeled as integrated Brownian motion. The
method described in \citet{Holt2013} was also tested in the SCEC
transient detection exercise. \citet{Holt2013} detect transient signal
from an estimate of transient strain rates, and they estimate
transient strain rates using a bi-cubic spatial parameterization of
displacements between time epochs. \citet{Holt2013} define a detection
threshold based on the transient strain rate magnitude, and we
demonstrate that this is indeed an effective criterion for identifying
a geophysical signal. For the same reasons described above, the
transient deformation and corresponding uncertainties estimated by
these methods can be biased by the chosen spatial parameterization. It
is then difficult to distinguish signal from noise with these methods,
which limits their utility for transient detection.

@@

Our outlier detection algorithm is loosely based on the data editing
algorithm from \citet{Acheson1975}. Let $\data^*$ denote all $n$ GNSS
displacement observations for a single directional component, which
have been made at positions and times $\points$.  We describe
$\data^*$ as a realization of the random vector
\begin{equation}\label{eq:OutlierDataVec}
\data = \G\mitbf{m} + v(\points) + \mitbf{w},
\end{equation}
where $\G$ and $\mitbf{m}$ are the same as in eq. (\ref{eq:DataVec}),
$v$ is a Gaussian process distributed as $\mathcal{GP}(0,C_v)$, and
$\mitbf{w}$ is a vector of uncorrelated Gaussian noise with known
standard deviations $\mitbf{\sigma} = [\sigma_1, \sigma_2, ...,
\sigma_n]$. The Gaussian process $v$ is intended to describe transient
features in the data that cannot be explained by the linear trend or
seasonal terms in $\G$. We let the temporal covariance of $v$ be a
squared exponential, and we let $v$ be spatially uncorrelated so that
\begin{equation}
C_v(p,p') = \phi^2\exp\left(\frac{-|t - t'|^2}{2\tau^2}\right) \delta_{\pos,\pos'},
\end{equation}
where $\delta_{\pos,\pos'} = 1$ if $\pos = \pos'$ and $0$ otherwise.
The spatial covariance of $v$ has little effect on the detected
outliers, and so we have assumed that $v$ is spatially uncorrelated
for simplicity. Based on our experience, $v$ can reasonably describe
most transient features in the data when we set $\phi = 1$ mm and
$\tau = 10$ days.
